---
title: "Loading and visualizing growth rate data"
layout: page
output: 
    md_document:
        variant: markdown
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(fig.cap="figure")
```

------------

> ### Learning Objectives
>
> * Use some of R's key functions for loading and understanding data sets
> * Explore the **`ggplot2`** package to visualize growth rate data


------------

To follow along without lots of copying and pasting, you can download 
[a script with nothing but the R code here](../../scripts/E-01-growth-rates.R).  

## Loading some growth rate data

In this exploration we'll load some real experimental growth rate data into R. If you 
haven't already, the readings [introducing R](../../readings/R-01-intro-to-r), 
[starting with data](../../readings/R-02-starting-with-data), and 
[using ggplot2](../../readings/R-04-visualization-ggplot2) provide some background
to the coding we'll use here.

In some cases we might start with data stored locally on our computer, or online 
via a link.  But here we'll take advantage of a data set that is built into an
R package called **`growthrates`**. Unless you've been doing some growth rate
modeling on your own, you probably don't have this package installed 
yet.  In RStudio, install it with the `install.packages()` function, calling `install.packages("growthrates")`.  Now let's load it into our R session with 
library().

```{r, eval=TRUE}
## only uncomment the next line if you need to install growthrates
#install.packages("growthrates")

## loading the package
library(growthrates)
```


## Getting to know the new package and data

Okay, so we've loaded the package. Let's use `?` to look at the help file.

```{r, eval=FALSE}
?growthrates
```

From the help file, you can see that this package has several functions to fit 
models to experimental growth rate data. From the examples at the bottom it looks
like they've included some data sets.  They work with the data `bactgrowth` in 
this example.  If you want to know what data are included in a package, the
`data()` function can help you.

```{r, eval=FALSE}
data(package = "growthrates")
```

So we can see that there are two datasets: *bactgrowth* and *antibiotic*.  It 
turns out that *bactgrowth* is an extremely messy dataset that the package authors
included to show how their model-fitting works on tricky data.  *antibiotic* is 
a bit easier to interpret.  Let's use the ? function to see what these data are.

```{r, eval=FALSE}
?antibiotic
```

We can see that these data are growth rates for the bacterium [Pseudomonas putida](https://en.wikipedia.org/wiki/Pseudomonas_putida), grown at different 
concentrations of the antibiotic tetracycline.  This is an interesting bacterium;
it may prove useful for clearing polluted soils of some byproducts of petroleum
refining.  Even though we've loaded the package **`growthrates`**, we still need to
load the data.  Here the `data()` function comes into play again.

```{r}
data(antibiotic)
```

> ### Advanced note -- the diverse outputs of a single function
> 
> You may be surprised that the `data()` function could yield a list of data, or 
> load a particular data set, depending on the arguments you enter. Many functions 
> in R can take a range of arguments, and the output many depend on what arguments
> you chose to specify.  The help function `?` (here you'd use `?data`) is useful for
> exploring everything that a function can do.

Let's get back to the data.  After running `data(antibiotic)` you can see in the 
upper right of RStudio that we have a new object in our globl environment. 
Now we must figure out how it is organized. There are several functions that 
summarize a data set for easy reading. One particularly useful one is `str()`.

```{r}
str(antibiotic)
```

What is this summary telling us?  A lot!  We see that there are 2928 observations
(rows) in our data, and each observation has 5 variables (columns).  `time` and 
`value` are our key variables here.  `time` tells us the time in hours from the
beginning of each experiment, and `value` tells us the bacterial concentration 
at that time point in that well.  This is all measured using optical density in
a plate reader -- there are many separate wells of media and bacteria that are 
all growing in parallel.  By measuring the optical density (absorbance) of the 
fluid to light, you can estimate the bacterial population size. In this case they 
didn't use a reference to translate these numbers back to a population size, but
the relative change should still be useful for seeing the pattern of growth. 
*(Note: this assumes that the optical density changes linearly with the bacterial
concentration.)*

There are a few other variables. `conc` tells us the concentration of 
tetracylcine used in each well, and `repl` tells us which replicate.  They had
four replicates.  In other words, they repeated each growth experiment with the 
exact same antibiotic concentration 4 times. This helps check that they are getting
consistent results.

> ### Challenge -- sources of noise
>
> * Based on the results from a single lab, would you be confident drawing
>   conclusions about the growth of this species in general?
> * What are some other sources of variation that might influence the growth rate data?
>   (To think about this, imagine all the decisions, explicit or implicit, made 
>   by the researchers as they set this up.)


## Visualizing our data

These data are already pretty tidy and well-organized for plotting. We'll use the
**`ggplot2`** package to make plots.  It's easy to work with, flexible, and can quickly 
make high quality data visualizations. Let's load it and get started. Instead of 
explaining everything, you can try to fiddle with the code to start getting a 
feel for the syntax.  If you've worked through [Reading 3: Manipulating and 
analyzing data with tidyverse](../../readings/R-03-dplyr), you've already installed
the **`tidyverse`** package, which includes **`ggplot2`**. If not, the code
`install.packages("ggplot2")` can install just that package.

```{r}
library(ggplot2)
ggplot(antibiotic,aes(x=time,y=value))+geom_point()
```

With one short line of code, we made a pretty nice plot. A few notes about what
I entered into the `ggplot()` function. I first entered the data frame (our matrix,
`antibiotic`).  Then I entered the *aesthetics* using `aes()`, 
specifying that the x-axis represents the variable *time* and the y-axis represents 
*value*.  These names need to exactly match the variable names from our data frame.  

So, it looks okay, but we've obscured the information about antibiotic concentration.
We can also specify color using `aes()`.  Let's represent the concentrations using
different colors.

```{r}
ggplot(antibiotic,aes(x=time,y=value,color=conc))+geom_point()
```

Hmmm, that doesn't look great.  The automatic scale doesn't work well when 
some values are very close together while others are spread apart.  We could 
dive into the nitty gritty of specifying custom color scales.  We could 
transform the data.  Or 
we could use a little hack -- it turns out that `ggplot()`
works better in these cases when the variable is a discrete (or factor) scale, not a 
continuous scale.  Don't stress if this feels confusing, but let's just try turning
the concentrations into a discrete factor variable instead.  *(If you need a refresher
on factors, [Reading 2: Starting with data](../../readings/R-02-starting-with-data) 
explains what a factor is and how R works with them.)*

```{r}
ggplot(antibiotic,aes(x=time,y=value,color=factor(conc)))+geom_point()
```

That's a bit easier to read.  Now we can clearly see the different replicates within
a single concentration, and we can distinguish pretty well between the concentration
gradients. As discussed in the [ggplot2 reading](../../readings/R-04-visualization-ggplot2),
we can also use *facets*, making an individual plot for each antibiotic concentration.
Data visualization experts sometimes call these *small multiples*, and they can 
be a nice way to view a lot of data at once.

```{r}
ggplot(antibiotic,aes(x=time,y=value)) +
    geom_point() +
    facet_wrap(~ conc)
```

> ### Challenge -- interpreting the data
>
> Let's interpret these graphs.
>
> * How is concentration affecting growth rates? Are there any outliers?
> * Is this growth linear? Exponential? How would you describe it?
> * What conclusions can you draw from these data?


```{r, child="../_page_built_on.Rmd", purl=FALSE}
```
