---
title: "Conditional probability"
layout: page
output: 
    md_document:
        variant: markdown
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(fig.cap="figure")
```

------------

> ### Learning Objectives
>
> * Build intuition for how the probability of some event can depend on other
>   evidence.
> * Be able to analyze a potential new test to screen for a disease
> * Use and interpret equations to calculate conditional probabilities

------------

As always, you can download [a script with nothing but the R code here](../../scripts/E-conditional-probability.R).    

## The challenge of screening for rare diseases

When a disease is common in a population, it can be a good use of public health 
funds to conduct *screening*. [Screening](https://www.hopkinsmedicine.org/healthlibrary/conditions/pathology/screening_tests_for_common_diseases_85,P00965) 
is the process of testing a focal population
with the aim of detecting a disease early. In some cases, this allows for treatment to
prevent disease progression. For example, the Pap test can detect 
cervical cancer at an early, asympomatic stage, leading to effective early treatment.
Another common example, a cholesterol blood test, can help inform lifestyle changes
or allow medical intervention.  

What exactly does this have to do with probability?  Well, it turns out the details
of the test, and the prevalence of the disease in the focal population you're testing, play a big role in how useful the test results really are.  To understand this, 
we'll need to define a few useful terms to describe test features. 

> * **Sensitivity** is the probability that a person who actually has the disease
>   will get a positive test result from the screening test.  This is also
>   called the **true positive rate**.
> * **Specificity** is the probability that a person who does not have the disease will
>   get a negative test result from the screening test.  This is also called the
>   **true negative rate**.

Both of these features vary widely, depending on the test. And they critically 
relate to the disease prevalence. To build intuition about this, open up [this
screening app](https://rfurrow.shinyapps.io/DiseaseScreening/) in a new tab.

Although we haven't directly mentioned it, these screening tests exemplify how
we think about *conditional probability*. Conditional probability refers to 
the probability of a event, ***given*** that some other event has occurred. We
could rephrase sensitivity as the probability of a positive test result, `+`, ***given***
that the person has the disease.  And specificity is the probability of a negative
test result, `-`, ***given*** that the person does not have the disease.  


```{r, child="../_page_built_on.Rmd"}
```
