---
title: "Modeling growth rates"
layout: page
output:
  md_document:
    variant: markdown_github
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(fig.cap="figure")
```

------------

> ### Learning Objectives
>
> * Work with growth rate data to explore mathematical growth models.
> * Manipulate model parameters to build intuition.

------------

As always, you can download [a script with nothing but the R code here](../scripts/E-02-growth-rate-models.R).  

## Loading the growth rate data.

You can remind yourself of these data by reviewing the [previous reading](../readings/E-01-growth-rates).  Let's quickly load these data again.

```{r, eval=TRUE}
library(growthrates)
data(antibiotic)
str(antibiotic)
```


Okay, here's the same quick summary of our data that we saw before.  Let's pull
up that final plot as well.  

```{r eval=TRUE}
library(ggplot2)
ggplot(antibiotic,aes(x=time,y=value,color=factor(conc)))+geom_point()
```

Wow, these look like some pretty well-behaved data!  The replicates at the same 
antibiotic concentration are pretty similar to each other, while bacteria grown 
at increasing antibiotic concentrations show decreasing growth rates.  Trying to
build a model to include time and antibiotic concentration is a bit above our paygrade
right now.  Let's just focus on a single concentration; why not start at 0. The package
`dplyr` has some nice tools for quickly filtering data to particular subsets. Don't
forget to install it first!

```{r eval=TRUE}
library(dplyr)
antibiotic_0 <- filter(antibiotic, conc == 0)
ggplot(antibiotic_0,aes(x=time,y=value))+geom_point()
```

Okay, we have a few things to unpack.  First, we got a suprise message when we 
loaded `dplyr`.  


```{r, child="../_page_built_on.Rmd"}
```
