---
title: "Modeling growth rates"
layout: page
output:
  md_document:
    variant: markdown
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(fig.cap="figure")
```

------------

> ### Learning Objectives
>
> * Work with growth rate data to explore mathematical growth models.
> * Manipulate model parameters to build intuition.

------------

As always, you can download [a script with nothing but the R code here](../scripts/E-02-growth-rate-models.R).  

## Loading the growth rate data

You can remind yourself of these data by reviewing the [previous reading](../readings/E-01-growth-rates).  Let's quickly load these data again.

```{r, eval=TRUE}
library(growthrates)
data(antibiotic)
str(antibiotic)
```


Okay, here's the same quick summary of our data that we saw before.  Let's pull
up that final plot as well.  

```{r eval=TRUE}
library(ggplot2)
ggplot(antibiotic,aes(x=time,y=value,color=factor(conc)))+geom_point()
```

Wow, these look like some pretty well-behaved data!  The replicates at the same 
antibiotic concentration are pretty similar to each other, while bacteria grown 
at increasing antibiotic concentrations show decreasing growth rates.  Trying to
build a model to include time and antibiotic concentration is a bit above our paygrade
right now.  Let's just focus on a single concentration; why not start at 0. The package
`dplyr` has some nice tools for quickly filtering data to particular subsets. Don't
forget to install it first!

```{r eval=TRUE}
library(dplyr)
antibiotic_0 <- filter(antibiotic, conc == 0)
ggplot(antibiotic_0,aes(x=time,y=value))+geom_point()
```

Okay, we have a few things to unpack.  First, we got a suprise message when we 
loaded `dplyr`.  It mentions some "objects" from other packages that are now "masked".
This is because dplyr has functions named filter, lag, and intersect, setdiff, etc.  But
functions with those names already exist in R.  So, when `dplyr` is loaded, it masks the 
other version.  Thus, when you call the function `filter()`, it will be the function 
from the `dplyr` package, not the basic function from the `stats` package. This is usually 
not a big deal, but it's something to remember if a function seems to be working oddly.

Second, we didn't define the *color* aesthetic this time.  Why not? Well, because 
we are only looking at a single concentration level, so we don't need it.

## Comparing with a theoretical model

Okay, so how do we model this? If you're currently taking BIS 23a or a similar course,
you may have learned about this in lecture.  For a refresher, this [Khan Academy page](https://www.khanacademy.org/science/biology/ecology/population-growth-and-regulation/a/exponential-logistic-growth)
has a walkthrough about modeling population growth.

The simplest model is often the best. If it fits the data well, then you can have
clear interpretations and predictions. Let's start with basic, exponential growth 
and see how it looks. Fundamentally, the exponential model is based on the assumption
that a population's growth rate is proportional the current population size. In other 
words, even though each individual reproduces at a similar rate, on average, the actual 
rate of new offspring will go up as the population increases.  For humans, there are 
many factors that make the global population growth different from exponential.  
[Our World in Data](https://ourworldindata.org/world-population-growth) has 
a nice deep dive into human population growth trends.

But for bacteria, particularly at small population sizes, the exponential model
might fit well. In mathematical terms, our assumption of proportional growth rate
can be translated into an equation as $\frac{dN}{dt} = rN$.  Here $N$ is the population
size, $t$ is the time variable, and $r$ is a parameter to represent the growth rate.
You can think of $r$ as a parameter that represents net growth, incorporating both births
and deaths.  If a population is decreasing, it's totally possible that $r$ could be 
negative, when births are occurring less often than deaths. To put the equation into words, the change in population as a particular moment equals the growth rate $r$ times the population size $N$.

This is not a course on differential equations. But it turns out that the solution 
to this equation is pleasantly simple: $N(t) = e^{rt}N(0)$. So the population at 
some time $t$ is a function of $N(0)$, the intial population, times $e^{rt}$.  So
$N(t)$ is an expontential function of $t$, hence we call this the exponential model.

But does it fit our data well? This next step will get you used to the typical way
that we visualize models in R.

##Visualizing the exponential model

To plot the function, we can generate model output by taking a vector of
time points, and applying the function to each one. It feels confusing at first,
but it's surprisingly simple to do. 

```{r,eval=TRUE}
antibiotic_0 <- mutate(antibiotic_0, model_output = exp(0.57*time)*0.01)
p <- ggplot(antibiotic_0,aes(x=time,y=value))+geom_point()
p <- p + geom_line(aes(y=model_output), color = "red")
p <- p + ylim(c(0,1))
p
```



```{r, child="../_page_built_on.Rmd"}
```
