---
title: "Before we start"
layout: page
output:
  md_document:
    variant: markdown_github
---


```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

------------

> ### Learning Objectives
>
> * Describe the purpose of the RStudio Script, Console, Environment, and Plots
>   panes.
> * Organize files and directories for a set of analyses as an R
>   Project, and understand the purpose of the working directory.
> * Use the built-in RStudio help interface to search for more information on R
>   functions.
>  * Demonstrate how to provide sufficient information for
>   troubleshooting with the R user community.

------------


## What is R? What is RStudio?

The term "`R`" is used to refer to both the programming language and the
software that interprets the scripts written using it.

[RStudio](https://rstudio.com) is currently a very popular way to not only write
your R scripts but also to interact with the R software. To function correctly,
RStudio needs R and therefore both need to be installed on your computer.


## Why learn R?

### R does not involve lots of pointing and clicking, and that's a good thing

The learning curve might be steeper than with other software, but with R, the
results of your analysis do not rely on remembering a succession of pointing
and clicking, but instead on a series of written commands, and that's a good
thing! So, if you want to redo your analysis because you collected more data,
you don't have to remember which button you clicked in which order to obtain
your results; you just have to run your script again.

Working with scripts makes the steps you used in your analysis clear, and the
code you write can be inspected by someone else who can give you feedback and
spot mistakes.

Working with scripts forces you to have a deeper understanding of what you are
doing, and facilitates your learning and comprehension of the methods you use.

### R code is great for reproducibility

Reproducibility is when someone else (including your future self) can obtain the
same results from the same dataset when using the same analysis.

R integrates with other tools to generate manuscripts from your code. If you
collect more data, or fix a mistake in your dataset, the figures and the
statistical tests in your manuscript are updated automatically.

An increasing number of journals and funding agencies expect analyses to be
reproducible, so knowing R will give you an edge with these requirements.


### R is interdisciplinary and extensible

With 10,000+ packages that can be installed to extend its capabilities, R
provides a framework that allows you to combine statistical approaches from many
scientific disciplines to best suit the analytical framework you need to analyze your
data. For instance, R has packages for image analysis, GIS, time series, population
genetics, and a lot more.


### R works on data of all shapes and sizes

The skills you learn with R scale easily with the size of your dataset. Whether
your dataset has hundreds or millions of lines, it won't make much difference to
you.

R is designed for data analysis. It comes with special data structures and data
types that make handling of missing data and statistical factors convenient.

R can connect to spreadsheets, databases, and many other data formats, on your
computer or on the web.


### R produces high-quality graphics

The plotting functionalities in R are endless, and allow you to adjust any
aspect of your graph to convey most effectively the message from your data.


### R has a large and welcoming community

Thousands of people use R daily. Many of them are willing to help you through
mailing lists and websites such as [Stack Overflow](https://stackoverflow.com/), or on the [RStudio community](https://community.rstudio.com/).


### Not only is R free, but it is also open-source and cross-platform

Anyone can inspect the source code to see how R works. Because of this
transparency, there is less chance for mistakes, and if you (or someone else)
find some, you can report and fix bugs.



## Knowing your way around RStudio

Let's start by learning about [RStudio](https://www.rstudio.com/), which is an
Integrated Development Environment (IDE) for working with R.

The RStudio IDE open-source product is free under the
[Affero General Public License (AGPL) v3](https://www.gnu.org/licenses/agpl-3.0.en.html).
The RStudio IDE is also available with a commercial license and priority email
support from RStudio, Inc.

We will use RStudio IDE to write code, navigate the files on our computer,
inspect the variables we are going to create, and visualize the plots we will
generate. RStudio can also be used for other things (e.g., version control,
developing packages, writing Shiny apps) that we will not cover during the
workshop.


```{r}
x <- 7
x
```